Demographic history and patterns of selection in maize during and since domestication

### 2015-07-31 manuscript.tex
Draft to edit

\title{Demographic history and patterns of selection in maize during and since domestication}
\author{Timothy M. Beissinger\affil{1}{Dept. of Plant Sciences, University of
    California, Davis, CA, USA}, Li Wang\affil{2}{Iowa State University, Ames, IA, USA}, Arun
  Durvasula\affil{1}{}, Kate Crosby\affil{1}{}, Matthew Hufford\affil{2}{}, \and Jeffrey
  Ross-Ibrarra\affil{1}{}\affil{3}{Genome Center and Center for population biology, University of
    California, Davis, CA, USA} }

\significancetext{Patterns of linked selection, or the impact of selection on sites that neighbor the functional variant, have been carefully evaluated in only a few species. In this work, we demonstrate that selection against deleterious mutations leaves a pronounced signature on the maize genome, reducing diversity in and immediately around genes. We show how demography interacts with selection to impact genome-wide patterns of diversity, including the important observation that rapid population expansion can increase the efficiency of selection as much as a sudden population bottleneck can weaken it. Along the way, we develop the first estimate the domegraphic parameters of the maize domestication from whole genome sequence data.}

@@

\begin{abstract}
  Unique selective and demographic forces operate on plants during domestication. These forces interact to generate the patterns of DNA variability that are persistent in crop species today. To quantify this interplay, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. We utilized whole genome sequence data from 23 maize and 13 teosinte individuals to make inferences.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize experienced a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps, specifically positive selection on new genic mutations, are not the primary force driving maize evolution. Also, we find that a reduced population size during domestication decreased the efficiency of purifying selection at purging deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels superior to those seen in teosinte. Our results demonstrate that in domesticated species, demography and selection are constantly fluctuating through time.
\end{abstract}

\dropcap{D}omesticated plant species evolve in a unique fashion
compared to their wild counterparts \cite{doebley2006}. This
is a result of both the anthropomorphic nature of artificial selection on
domesticates \cite{purugganan2009} as well as the demographic characteristics of the domestication
bottleneck(s) that they tend to have experienced
\cite{ross2007}. However, the
complex interplay between selective pressures and demographic
limitations, and the impact that this interplay has on identifying
selection and understanding demography, is not fully understood. Although a large body of
research that involves searching the genomes of domesticated species for evidence
of positive selection exists \cite{hufford2012, he2011, vigouroux2002, chapman2008}, these studies tend to focus on
identifying or mapping particular genes or regions that play an
important role in phenotypic evolution. In contrast, knowledge regarding the impacts that demography and
selection have on whole-genome patterns of genetic variability therefore remains limited.

@@

To that end, the objectives of our study were to 1) investigate the
relative importance of different forms of selection on whole-genome
variability in both maize and teosinte, as different forms of selection affect DNA
variability remarkably differently; 2) research the impact that the
domestication process has had on genetic variability in maize, and how
this compares to the impact of a different demographic history in
teosinte; and 3) precisely estimate the parameters of the
maize domestication bottleneck. We show that our third objective, estimating the
parameters of domestication, is not possible without first completing
 objectives one and two, which demonstrate that as in humans \cite{gazave2014}, the
majority of maize non-genic DNA may reasonably be treated as
neutral. We achieve these objectives by utilizing
whole-genome-sequence information from 23 maize and 13 teosinte lines
sequenced as part of the Maize HapMap 2 panel \cite{chia2012}.

\section{Results}
\subsection{Patterns of diversity differ between genic and
  nongenic regions of the genome}
Previous research of the maize domestication process has relied upon
observations drawn from genic DNA \cite{wright2005,wang1999,eyre1998}. Our data were generated through whole genome sequencing,
which eliminated this constraint. Importantly, we observed substantial
differences in patterns of diversity between genic and non-genic regions
of the genome for both maize and teosinte. For maize, mean pairwise
diversity ($\pi$) within genes was significantly lower than at
positions at least 5kb away from genes (0.00668 within, 0.00691 away, p$<$2e-44). The same
pattern of significantly greater diversity outside of genes was observed in
teosinte, but to a larger extent. For teosinte we observed $\pi$
within genes to be 0.0088 and at least 5kb from genes it was 0.0115
(p$\approx$0). These observations suggest that genes are not evolving
neutrally in maize or teosinte. Instead, some form of selection is likely
reducing diversity within genes. Additionally, these observations
suggest that demographic inference, which relies on the assumption of
neutrally evolving DNA, will be more accurate if it is based on
observations from non-genic genetic material.

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly 1-1.2 million years before
present \cite{ross2009}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). This observation suggests that hard sweeps are not a primary form of
selection for either maize or teosinte.

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene tga1, one of the few known instances of a hard sweep corresponding to a causal nonsynonymous mutuation in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causitive substitution to levels much lower than seen for synonymous substitutions (Figure S1), confirming that this method should be informative if hard sweeps were the prevaling mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites (Figure S2).

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_June}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_June}
\caption{Pairwise diversity surrounding synonymous and nonsynonymous
  substitutions in maize and teosinte. {\bf A:} Maize diversity; {\bf B:} Teosinte
diversity. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

\subsection{Pairwise diversity is heavily influenced by purifying selection}
% \jri{we will need to define background selection and be careful about ``purifying selection'' vs ``background selection'', as technically background selection does not include very weak deleterious alleles.}
Purifying selection refers to the situation where deleterious mutations arising in a population are continuously selected against.
When this form of selection is operating, it can serve to reduce genetic variability at linked neutral sites, a phenomenon often called background selection \cite{charlesworth1993}. Purifying and background selection lead to lower diversity within genes and other functional sites relative to neutral regions \textcolor{red}{(ref here)}.
We investigated purifying selection in maize and teosinte by evaluating the average magnitude of reduced diversity within genes and recovery away from genes in both taxa.
To enable comparisons between maize and teosinte, we standardized by neutral levels within each taxa, where neutral levels were defined as mean diversity at positions far from genes at least 0.01 cM distal from genes. For both maize and teosinte, pairwise diversity was reduced within genes and a recovery of diversity away from genes was observed. Interestingly, a stronger reduction of diversity and slower recovery was observed for teosinte than for maize, implying that purifying selection has left a more pronounced signature on pairwise diversity in the teosinte genome (Figure \ref{fig:purify}).

\begin{figure*}[t]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a smaller range on the x-axis. \label{fig:purify}
  %\jri{should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.''}
  }
\end{figure*}

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012} \textcolor{red}{(supplemental table here)}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\cite{tajima1989} was therefore shifted
to more positive values \ref{fig:tajima1989}. This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. This observation is consistent with previous research \cite{hufford2012}. Therefore, for demographic modeling we restricted analysis
to sites at least 5kb from gene start/stop postions.

\begin{figure}[b]
\centering
\includegraphics[width=.5\textwidth]{FigsAndFiles/SFS_and_Tajima.png}
\caption{The folded SFS and Tajimas D were calculated both inside and outside of genes for maize and teosinte. In both taxa, a dearth of rare alleles was observed within genes relative to outside of genes, which led to higher values of Tajima's D outside of genes than inside.{\bf A:} Folded SFS for maize; {\bf B:} Folded SFS for teosinte; {\bf C:} Maize Tajima's D; {\bf D:} Teosinte Tajima's D.  \label{fig:sfs} }
\end{figure}

We used diffusion approximation as implemented in
dadi \cite{gutenkunst2009} to find the domestication parameters that best explain
the joint site frequency spectrum of maize and teosinte.  The model we optimized began with an
equilibrium ancestral population of size $N_a$
splitting into separate maize and teosinte populations $T_B$ generations in the past. Moving forward in
time from the split, teosinte maintained the ancestral population size, while
maize experienced an immediate effective population size change to
$N_b$ individuals, followed by exponential growth to size
$N_m$. Additionally, since the split a $M_{tm}$ individuals migrated
from teosinte into maize and $M_{mt}$ individuals migrated from maize
to teosinte.

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) per base pair of
0.014734, which approximately matches a previous independent estimate \cite{eyre1998}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \cite{clark2005}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{fig:bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data which suggests maize domestication began
no more than 10,000 years before present \cite{smith1995}. This could reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate.

We utilized a complementary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was utilized since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate $N_e =
  992,713.5$. Although less biased than the previously mentioned
estimate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

\begin{figure}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameters of domestication as estimated by dadi. Notably,
  the maize effective population size ($N_e$) during the domestication
  bottleneck appears to have consisted of approximately 5.26\%
  the ancestral $N_e$, before recovering two at least 2.98
  times as large as the ancestral $N_e$. \label{fig:bottleneck} }
\end{figure}

\section{Post-domestiction expansion is apparent in patterns of singleton diversity}
Motivated by the rapid post-domestication expansion in maize is seen in our demographic analysis, we conducted an analysis of genome-wide patterns of singleton diversity. Singleton diversity measures the abundance of alleles that are only observed in one individual among the sample. As a class, singleton alleles depict the most recent patterns of evolution, but
they also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged compared to what we saw with respect to pairwise diversity. Maize singleton diversity was at least as reduced as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been more influenced by purifying selection than has
teosinte.

@@

Similarly, to complement our investigation of pairwise diversity surrounding substitutions, we studied patterns of singleton diversity surrounding synonymous and nonsynonymous substitutions in maize (Figure S3). A nearly identical pattern to that shown in Figure \ref{fig:hardSweeps} was observed. This further demonstrates that even the most recent selection patterns in maize are not dominated by hard sweeps.

\subsection{Purifying seleciton results are robust to soft-sweeps}
As a category, instances of soft-sweeps, or cases of selection on alleles that have already reached intermediate frequency before becoming advantageous, are difficult to identify \textcolor{red}{(refs here)}. However, this form of selection is likely to be frequent during crop domestication \textcolor{red}{(refs here)}. We were concerned that the disparate patterns of pairwise and singleton diversity in maize and teosinte, which seem to reflect differences in the efficiency of purifying selection due to demographic history, may instead reflect differences in the location and abundance of soft-sweeps. Therefore, we re-analyzed patterns of pairwise and singleton diversity in maize and teosinte including only a subset of sites that are least likely to correspond to soft-sweeps. To achieve this, we performed a genome-wide scan for soft or hard sweeps in maize or teosinte based on the H12 statistic \cite{garud2015}. H12 is sensitive to both hard and soft sweeps. Sites nearest to genes that displayed among the top 20\% of H12 values in either maize or teosinte were removed from our dataset, and both pairwise and singleton diversity at remaining sites, as a function of the distance to the nearest gene, was calculated (Figure S4). This subset of the data in no longer subject to the potentially confounding effects of soft-sweeps, and nearly identical patterns of pairwise and singleton diversity surrounding genes are observed. This provides further evidence that the patterns of diversity we observe in maize as compared to teosinte are the result of rapid population expansion as opposed to differing selective patterns in the domesticated and wild taxa.

\section{Discussion}
\subsection{Little positive selection on new genic mutations}
Our findings indicate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize. More specifically, we observe that positive selection on new genic mutations in maize is not common compared to other drivers of diversity. This finding does not exclude the possibility that hard sweeps have taken place at non-genic sites, such as in those in enhancer regions. One known example of positive selection on a non-genic mutation involves the \emph{tb1} locus of maize, which is one of the best characterized examples of positive selection on a ``domestication gene'' in any crop \cite{clark2006}. However, the maize \emph{tb1} allele was already present in teosinte before domestication \cite{studer2011}, so this even this example demonstrates supports our finding that hard sweeps are rare. The \emph{gt1} locus is another well-characterized case of positive selection operating on standing variation at an enhancer region \cite{wills2013}. Instances of selection on standing variation such as these, often called soft sweeps, may be a major contributor to maize patterns of diversity \cite{beissinger2014}, and this could be part of the explanation as to why hard sweeps appear to be so rare, despite obvious morphological differences between maize and teosinte.

>>>>

\title{Demographic history and patterns of selection in maize during and since domestication}
\author{Timothy M. Beissinger\affil{1}{Dept. of Plant Sciences, University of
    California, Davis, CA, USA}\affil{2}{US Department of Agriculture, Agricultural Research Service, Columbia, MO, USA}\affil{3}{Division of Plant Sciences, University of Missouri, Columbia, MO, USA} Li Wang\affil{4}{Iowa State University, Ames, IA, USA}, Kate Crosby\affil{1}{}, Arun
  Durvasula\affil{1}{}, Matthew Hufford\affil{4}{}, \and Jeffrey
  Ross-Ibrarra\affil{1}{}\affil{5}{Genome Center and Center for population biology, University of
    California, Davis, CA, USA} }

\significancetext{Patterns of linked selection, or the impact of selection on sites that neighbor a functional variant, have been carefully evaluated in only a few species. In this work, we demonstrate that selection against deleterious mutations leaves a pronounced signature on the maize genome, reducing diversity in and immediately around genes. We show how demography interacts with selection to impact genome-wide patterns of diversity, including the important observation that rapid population expansion can increase the efficiency of selection as much as a sudden population bottleneck can weaken it. Along the way, we develop the first estimate the demographic parameters of the maize domestication from whole genome sequence data.}

@@

\begin{abstract}
  Unique selective and demographic operate on domesticated plant species. These forces interact during and after domestication to generate the patterns of DNA variability that are persistent in crop species today. To quantify the interplay between demography and selection, we investigated genetic diversity in maize, one of the most important crops for food, feed, and fuel world-wide. We utilized whole genome sequence data from 23 maize and 13 teosinte individuals to make inferences.  We obtained a complete estimate of the population size fluctuations and other demographic parameters experienced by maize as it was domesticated from teosinte. Here, we show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid population size expansion post-domestication. We observe that hard sweeps, specifically positive selection on new genic mutations, are not the primary force driving maize evolution. We find that a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize. However, expansion after domestication has since increased the efficiency of purifying selection to levels superior to those seen in teosinte. Our results demonstrate that in domesticated species, demographic and selective history in the ancient and recent past both contribute to genetic variability that is present today, providing substantial implications for the continued improvement of domesticated species.

\end{abstract}

\dropcap{D}omesticated plant species evolve in a unique fashion
compared to their wild counterparts \cite{doebley2006}. This
is a result of both the anthropomorphic nature of artificial selection on
domesticates \cite{purugganan2009} as well as the demographic characteristics of the domestication
bottleneck(s) that they tend to have experienced
\cite{ross2007}. However, the
complex interplay between selective pressures and demographic
limitations, and the impact that this interplay has on identifying
selection and understanding demography, is not fully understood. Although a large body of
research that involves searching the genomes of domesticated species for evidence
of positive selection exists \cite{hufford2012, he2011, vigouroux2002, chapman2008}, these studies tend to focus on
identifying or mapping particular genes or regions that play an
important role in phenotypic evolution. In contrast, knowledge regarding the impacts that demography and
selection have on whole-genome patterns of genetic variability remains limited.

@@

To that end, the objectives of our study were to 1) investigate the
relative importance of different forms of selection on whole-genome
variability in both maize and teosinte 2) research the impact that the
domestication process has had on genetic variability in maize, and how
this compares to the impact of a different demographic history in
teosinte; and 3) precisely estimate the parameters of the
maize domestication bottleneck. We show that our third objective, estimating the
parameters of domestication, is not possible without first completing
 objectives one and two, which demonstrate that as in humans \cite{gazave2014}, the
majority of maize non-genic DNA may reasonably be treated as
neutral. We achieve these objectives by utilizing
whole-genome-sequence information from 23 maize and 13 teosinte lines
sequenced as part of the Maize HapMap 2 panel \cite{chia2012}.

\section{Results}
\subsection{Patterns of diversity differ between genic and
  non-genic regions of the genome}
Previous research of the maize domestication process has relied upon
observations drawn from genic DNA \cite{wright2005,wang1999,eyre1998}. Our data were generated through whole genome sequencing,
which eliminated this constraint. Importantly, we observed substantial
differences in patterns of diversity between genic and non-genic regions
of the genome for both maize and teosinte. For maize, mean pairwise
diversity ($\pi$) within genes was significantly lower than at
positions at least 5 kb away from genes (0.00668 within, 0.00691 away, p$<$2e-44). The same
pattern of significantly greater diversity outside of genes was observed in
teosinte, but to a larger extent. For teosinte we observed $\pi$
within genes to be 0.0088 and at least 5 kb from genes it was 0.0115
(p$\approx$0). These observations suggest that genes are not evolving
neutrally in maize or teosinte. Instead, some form of selection is likely
reducing diversity within genes. Additionally, these observations
suggest that demographic inference, which relies on the assumption of
neutrally evolving DNA, will be more accurate if it is based on
observations from non-genic genetic material.

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous (specifically missense) and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly 1-1.2 million years before
present \cite{ross2009}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). This observation suggests that hard sweeps are not a primary form of
selection for either maize or teosinte.

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene \emph{tga1}, one of the few known instances of a hard sweep corresponding to a causal missense substitution in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causative substitution to levels much lower than seen for synonymous substitutions (Figure S1), confirming that this method should be informative if hard sweeps were the prevailing mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites (Figure S2).

\begin{figure*}[!htb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_June}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_June}
\caption{Pairwise diversity surrounding synonymous and non-synonymous
  substitutions in maize and teosinte. {\bf A:} Maize diversity; {\bf B:} Teosinte
diversity. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

\subsection{Pairwise diversity is heavily influenced by purifying selection}
% \jri{we will need to define background selection and be careful about ``purifying selection'' vs ``background selection'', as technically background selection does not include very weak deleterious alleles.}
Purifying selection refers to the situation where deleterious mutations arising in a population are continuously selected against.
When this form of selection is operating, it can serve to reduce genetic variability at linked neutral sites, a phenomenon often called background selection \cite{charlesworth1993}. Purifying and background selection lead to lower diversity within genes and other functional sites relative to neutral regions.
We investigated purifying selection in maize and teosinte by evaluating the average magnitude of reduced diversity within genes and recovery away from genes in both taxa.
To enable comparisons between maize and teosinte, we standardized by neutral levels within each taxa, where neutral levels were defined as mean diversity at positions at least 0.01 cM distal from genes. For both maize and teosinte, pairwise diversity was reduced within genes and a recovery of diversity away from genes was observed. Interestingly, a stronger reduction of diversity and slower recovery was observed for teosinte than for maize, implying that purifying selection has left a more pronounced signature on pairwise diversity in the teosinte genome (Figure \ref{fig:purify}).

\begin{figure*}[!htb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. Bootstrap-based 95\% confidence intervals are depicted via shading. Inset plots depict a smaller range on the x-axis. \label{fig:purify}
  %\jri{should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.''}
  }
\end{figure*}

\subsection{Demography of maize domestication}
To explore whether differences in purifying selection between maize
and teosinte can be explained by demographic processes, we
first estimated the parameters of maize domestication using large-scale
sequencing data involving 23 inbred maize
landraces and 13 teosinte inbred lines included in the HapMap 2 panel
\cite{chia2012}. The maize
lines were collected from across the Americas and the teosinte lines
came from central Mexico. Before estimating demography, we compared
the site frequency spectrum (SFS) in genic and non-genic regions, and
observed substantial differences in the evolution of these classes of
sites. For both maize and teosinte, the SFS within genes showed a
dearth of low-frequency alleles and Tajima's D
\cite{tajima1989} was therefore shifted
to more positive values, as previously shown (Figure \ref{fig:purify}). This is consistent with the aforementioned purifying
selection and indicates that genic regions are not evolving
neutrally. Therefore, for demographic modeling we restricted analysis
to sites at least 5 kb from gene start/stop positions.

\begin{figure}[!htb]
\centering
\includegraphics[width=.5\textwidth]{FigsAndFiles/SFS_and_Tajima.png}
\caption{The folded SFS and Tajimas D were calculated both inside and outside of genes for maize and teosinte. In both taxa, a dearth of rare alleles was observed within genes relative to outside of genes, which led to higher values of Tajima's D outside of genes than inside.{\bf A:} Folded SFS for maize; {\bf B:} Folded SFS for teosinte; {\bf C:} Maize Tajima's D; {\bf D:} Teosinte Tajima's D.  \label{fig:sfs} }
\end{figure}

We used diffusion approximation as implemented in
dadi \cite{gutenkunst2009} to find the domestication parameters that best explain
the joint site frequency spectrum of maize and teosinte.  The model we optimized began with an
equilibrium ancestral population of size $N_a$
splitting into separate maize and teosinte populations $T_B$ generations in the past. Moving forward in
time from the split, teosinte maintained the ancestral population size, while
maize experienced an immediate effective population size change to
$N_b$ individuals, followed by exponential growth to size
$N_m$. Additionally, since the split $M_{tm}$ individuals migrated
from teosinte into maize and $M_{mt}$ individuals migrated from maize
to teosinte.

The most likely model suggested an ancestral $\theta$ ($4N\mu$) per base pair of
0.014734, which approximately matches a previous independent estimate \cite{eyre1998}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \cite{clark2005}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{fig:bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data, which suggest maize domestication began
no more than 10,000 years before present \cite{smith1995}. This could reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede physiological changes that could be identified
archaelogically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since our limited sample size provides low
power to detect rare alleles, we expect that the estimate of  $N_m = 366,973$, or $\sim 3N_a$, is likely an underestimate.

Therefore, we utilized a complementary data set of 4,021 maize land-race individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was utilized since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate $N_e =
  992,713.5$. Although less biased than the previously mentioned
estimate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

\begin{figure}[!htb]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameters of domestication as estimated by dadi. Notably,
  the maize effective population size ($N_e$) during the domestication
  bottleneck appears to have consisted of approximately 5.26\%
  the ancestral $N_e$, before recovering two at least 2.98
  times as large as the ancestral $N_e$. \label{fig:bottleneck} }
\end{figure}

\section{Post-domestication expansion is apparent in patterns of singleton diversity}
Motivated by the rapid post-domestication expansion in maize that is evident from our demographic analysis, we conducted an analysis of genome-wide patterns of singleton diversity. Singleton diversity measures the abundance of alleles that are only observed in one individual among the sample. As a class, singleton alleles depict the most recent patterns of evolution, but
they also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosinte (n=13), maize singletons were
analyzed directly as well as down-sampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was defined as diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged compared to what we saw with respect to pairwise diversity. Maize singleton diversity was at least as reduced as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been more influenced by purifying selection than has
teosinte.

@@

Similarly, to complement our investigation of pairwise diversity surrounding substitutions, we studied patterns of singleton diversity surrounding synonymous and non-synonymous substitutions in maize (Figure S3). A nearly identical pattern to that shown in Figure \ref{fig:hardSweeps} was observed. This further demonstrates that even the most recent selection patterns in maize are not dominated by hard sweeps.

\subsection{Purifying selection results are robust to soft-sweeps}
As a category, instances of soft-sweeps, or cases of selection on alleles that have already reached intermediate frequency before becoming advantageous, are difficult to identify \cite{wilson2014}. However, this form of selection is likely to be frequent during crop domestication \cite{innan2004}. We were concerned that the disparate patterns of pairwise and singleton diversity in maize and teosinte, which seem to reflect differences in the efficiency of purifying selection due to demographic history, may instead reflect differences in the location and abundance of soft-sweeps. Therefore, we re-analyzed patterns of pairwise and singleton diversity in maize and teosinte including only a subset of sites that are least likely to correspond to soft-sweeps. To achieve this, we performed a genome-wide scan for soft or hard sweeps in maize or teosinte based on the H12 statistic \cite{garud2015}. H12 is sensitive to both hard and soft sweeps. Sites nearest to genes that displayed among the top 20\% of H12 values in either maize or teosinte were removed from our data set, and both pairwise and singleton diversity at remaining sites, as a function of the distance to the nearest gene, was calculated (Figure S4). This subset of the data is no longer subject to the potentially confounding effects of soft-sweeps, and nearly identical patterns of pairwise and singleton diversity surrounding genes are observed. This provides further evidence that the patterns of diversity we observe in maize as compared to teosinte are the result of rapid population expansion as opposed to differing selective patterns in the domesticated and wild taxa.

\section{Discussion}
\subsection{Little positive selection on new genic mutations}
Our findings indicate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize. More specifically, we observe that positive selection on new genic mutations in maize is not common compared to other drivers of diversity. This finding does not exclude the possibility that hard sweeps have taken place at non-genic sites, such as in those in enhancer regions. One known example of positive selection on a non-genic mutation involves the \emph{tb1} locus of maize, which is one of the best characterized examples of positive selection on a ``domestication gene'' in any crop \cite{clark2006}. However, the maize \emph{tb1} allele was already present in teosinte before domestication \cite{studer2011}, so this this example is in agreement with our finding that hard sweeps are rare. The \emph{gt1} locus is another well-characterized case of positive selection operating on standing variation at an enhancer region \cite{wills2013}. Instances of selection on standing variation such as these, often called soft sweeps, may be a major contributor to maize patterns of diversity \cite{beissinger2014}, and this could be part of the explanation as to why hard sweeps appear to be so rare, despite obvious morphological differences between maize and teosinte.
